---
title: "Review of Surrogate Variable Analysis"
author: "Meilei Jiang"
data: July 22, 2015
output: html_document
---

## Data variation and heterogeneity

Data variation dues to different kinds of factors:

- modeled factors (primary variables): e.g. class labels
- unmodeled factors : e.g. batch effects:  age, sex
- other variation : e.g. white noise, systematic error

In practice, we want to estimate the effects of primary variables as well as discover the unmodeled factors. For instance, we want to training a model by given class labels. However, the effects of primary variables can be blurred by unmodeled factors. Even the data from the same class can have different distributions or a mixture of distribution, which is also called heterogeneity. 

The surrogate variable analysis (SVA) builds up a linear model for data variation and tries to construct surrogate variables to represent the batch effects.

### Models for data variation

The $p \times n$ random data matrix $X =(x_1, \dots, x_p)'$ contains $p$ features and each feature $x_i$ = $(x_{i1}, \dots, x_{in})'$ has the expression from $n$ samples. The primary variables are $Y = (y_1, \dots, y_n)'$. 

Assume that $\mathbb{E}[x_i|Y] = b_i S(Y) = b_i S$, where $S$ is a matrix of basis function of primary variables with dimension $d \times n$. Also, assume that the marginal model for each $e_i = x_i - b_i S$ is known or approximated sufficiently. In the matrix form, the model can be written as:
                                        $$X = B S + E$$
Moreover, we assume the following decomposition:
                                   $$X = B S + \Gamma G + U $$
Leek and Storey (2008) showed this decomposition exist in the following sense. Suppose that for each $e_i$, there is no Borel measurable function g such that $e_i = g(e_1, \dots, e_{i - 1}, e_{i + 1}, \dots, e_m)$ almost surely. Then there exist matries $\Gamma_{p \times m } G_{r \times n} (r <= n)$ and $U_{p \times n}$ such that:
                                        $$X = B S + \Gamma G + U$$
where the rows of $U$ are jointly independent random vectors so that
$$\mathbb{P}(u_1, u_2, \dots, u_m) = \mathbb{P}(u_1) \times \mathbb{P}(u_2) \times \dots \times \mathbb{P}(u_m)$$
Also $\forall i = 1, \dots, m, u_i \neq 0$ and $u_i = h_i(e_i)$ for a non-random Borel measurable function $h_i$. 


When the structure of $G$ is unknown, a set of surrogate variables $\hat{G}$ are constructed by SVA algorithm.

### Detect unmodeled factors

If there are no batch effects, i.e.$\Gamma = 0$,  the residual matrix $R = X - \hat{B} S$ should not have unusally large variations in some specific directions. To estimate the dimension $r$ of unmodeled factors, the number of unusally large eigenvalues need to be determined. 

In the SVA, an permutation test, a non-parametric approach, is proposed on the residual matrix $R$. It compares the proportion of singular values in $R$ to the proportion of singular values in randomized residual matrices, where each row is permuted individually to break down any structure across rows.

### Construct surrogate variables

The sva algorithm does not attempt to estimate the unmodeled factors $G$. In fact, the sva constructs surrogate variables to represent the variation not caused by primary variables. 

Say in the previous step, $\hat{K}$ unusally large eigenvalues are identified. For $\lambda_k, k = 1, \cdots, \hat{K}$, take out its corresponding right singular vectors $e_k$. In some sense, $e_k$ can act as surrogate variable $\hat{g}_k$. However, when $\hat{G}$ and $S$ are not orthogonal，$\hat{B}$, the estimation of the effects of $S$, is biased. So are $R$ and $e_k$.

To reduce this bias, next, subset of feasures $X^{(r)}$ which have a strong assosiation with $e_k$ are idtentified. Then refit the model with $X^{r}$ and obtain the residual matrix $R^{r}$. Last, identify the eigenvector $e_{j_k}^{r}$ of $R^{r}$ which has the largest correlation coefficient with $e_k$.

We can choose $e_{j_k}^{r}$ as $\hat{g}_k$ or continue this iteration. This approach can up-weight the features that show strong assotiation with $G$ and down-weight feasures that show strong association with $S$.

### Simulation study

The current sva algorithm requires that the data has at least 3 feasures. We consider a 2-class toy example and sample in each class has a Gaussian Mixture distribution. 

#### 3 feasures data set
Training data are divided into two class:

- Class A: $Y1 \sim (1 - \pi_1) \mathrm{N}\Big( (4, -4, 0)', I_3 \Big) + \pi_1 \mathrm{N}\Big( (4, 4, 0)', I_3 \Big)$, where $\pi_1 \sim \text{Binom}(0.5)$;
- Class B: $Y2 \sim (1 - \pi_2) \mathrm{N}\Big( (-4, -4, 0)', I_3 \Big) + \pi_2 \mathrm{N}\Big( (-4, 4, 0)', I_3 \Big)$, where $\pi_2 \sim \text{Binom}(0.5)$.

The plots of data are as following:

- $X_1$ is the primary variables
- $X_2$ is the batch effects
- $X_3$ is the random noise

```{r qplot, echo = FALSE, warning=FALSE, message = FALSE}
library(ggplot2)
library(reshape2)
library(dplyr)
load("/Users/meileijiang/researchspace/robust-against-heterogeneity/Surrogate Variable Analysis/rdata/toyexample.RData")
M3d <- train3d[,2:4]
# Expression data
Edata3d = t(M3d)
Mdata3d = melt(Edata3d)

g1 = ggplot(Mdata3d, aes(x = Var2, y = Var1, fill = value)) +
  labs(x = "Sample", y = "Gene", fill = "Value") +
  geom_tile() + 
  scale_fill_gradient2()
print(g1)
g2 = ggplot(data = train3d, aes(x = X1, y = X2, color = y)) + geom_point(aes(shape = batch))
print(g2)

```
Then, we apply sva algorithm to detect the number of surrogate variables.
```{r, echo = FALSE, warning=FALSE, message = FALSE}
library(sva)
# make model matrix -------------------------------------------------------
## full model 
mod_3d = model.matrix(~ y, data = train3d)
## null model
mod0_3d = model.matrix(~ 1, data = train3d)

# estimate the number of latent factors that need to be estimated ---------
n.sv_be_3d = num.sv(Edata3d,mod_3d,method="be")
print(n.sv_be_3d)

```

The algorithm fails to detect the number of SV.

#### 10 feasures data set
Training data are divided into two class:

- Class A: $Y1 \sim (1 - \pi_1) \mathrm{N}\Big( (2, -2, 0, \dots, 0)', I_{10} \Big) + \pi_1 \mathrm{N}\Big( (2, 2, 0, \dots, 0)', I_{10} \Big)$, where $\pi_1 \sim \text{Binom}(0.5)$;
- Class B: $Y2 \sim (1 - \pi_2) \mathrm{N}\Big( (-2, -2, 0, \dots, 0)', I_{10} \Big) + \pi_2 \mathrm{N}\Big( (-2, 2, 0, \dots, 0)', I_{10} \Big)$, where $\pi_2 \sim \text{Binom}(0.5)$.

The plots of data are as following:

- $X_1$ is the primary variables
- $X_2$ is the batch effects
- $X_3, \dots, X_10$ is the random noise

```{r, echo = FALSE, warning=FALSE, message = FALSE}
library(ggplot2)
library(reshape2)
library(dplyr)

M10d <- train10d %>% select(-y, -batch)
# Expression data
Edata10d = t(M10d)
Mdata10d = melt(Edata10d)

g4 = ggplot(Mdata10d, aes(x = Var2, y = Var1, fill = value)) +
  labs(x = "Sample", y = "Gene", fill = "Value") +
  geom_tile() + 
  scale_fill_gradient2()
print(g4)

```
Then, we apply sva algorithm to detect the number of surrogate variables.
```{r, echo = FALSE, warning=FALSE, message = FALSE}
library(sva)
# make model matrix -------------------------------------------------------
## full model 
mod_10d = model.matrix(~ y, data = train10d)
## null model
mod0_10d = model.matrix(~ 1, data = train10d)

# estimate the number of latent factors that need to be estimated ---------
n.sv_be_10d = num.sv(Edata10d,mod_10d,method="be")
print(n.sv_be_10d)

```

The algorithm fails to detect the heterogeneity.

#### 100 feasures data set
Training data are divided into two class:

- Class A: $Y1 \sim (1 - \pi_1) \mathrm{N}\Big( (2, -2, 0, \dots, 0)', I_{100} \Big) + \pi_1 \mathrm{N}\Big( (2, 2, 0, \dots, 0)', I_{100} \Big)$, where $\pi_1 \sim \text{Binom}(0.5)$;
- Class B: $Y2 \sim (1 - \pi_2) \mathrm{N}\Big( (-2, -2, 0, \dots, 0)', I_{100} \Big) + \pi_2 \mathrm{N}\Big( (-2, 2, 0, \dots, 0)', I_{100} \Big)$, where $\pi_2 \sim \text{Binom}(0.5)$.

The plots of data are as following:

- $X_1$ is the primary variables
- $X_2$ is the batch effects
- $X_3, \dots, X_100$ is the random noise

```{r, echo = FALSE, warning=FALSE, message = FALSE}
library(ggplot2)
library(reshape2)
library(dplyr)

M100d <- train100d %>% select(-y, -batch)
# Expression data
Edata100d = t(M100d)
Mdata100d = melt(Edata100d)

g5 = ggplot(Mdata100d, aes(x = Var2, y = Var1, fill = value)) +
  labs(x = "Sample", y = "Gene", fill = "Value") +
  geom_tile() + 
  scale_fill_gradient2()
print(g5)

Mdata100d.0 = melt(Edata100d[1:10, ])
g5.0 = ggplot(Mdata100d.0, aes(x = Var2, y = Var1, fill = value)) +
  labs(x = "Sample", y = "Gene", fill = "Value") +
  geom_tile() + 
  scale_fill_gradient2()
print(g5.0)
```

Then, we apply sva algorithm to detect the number of surrogate variables.
```{r, echo = FALSE, warning=FALSE, message = FALSE}
library(sva)
# make model matrix -------------------------------------------------------
## full model 
mod_100d = model.matrix(~ y, data = train100d)
## null model
mod0_100d = model.matrix(~ 1, data = train100d)

# estimate the number of latent factors that need to be estimated ---------
n.sv_be_100d = num.sv(Edata100d,mod_100d,method="be")
print(n.sv_be_100d)

```
Then we want to see the details of detecting the surrogate variables. Especially, we want to understand the scree plot in each permuation residual matrix.
```{r, echo = FALSE, warning=FALSE, message = FALSE}

library(dplyr)
library(ggplot2)
source("/Users/meileijiang/researchspace/robust-against-heterogeneity/Surrogate Variable Analysis/pcafuns/R/getPcaResult.R")
M <- train100d %>% select(-y, -batch)
# Expression data
Edata = t(M)
colnames(Edata) = paste0("sample", c(1: dim(Edata)[2]))
Mdata = melt(Edata)

# primary variable of interest
Y <- train100d %>% select(y)

# generate the basis matrix -----------------------------------------------

## full model 
mod = model.matrix(~ y, data = train100d)
## null model
mod0 = model.matrix(~ 1, data = train100d)

HatB = Edata %*% mod %*% solve(t(mod) %*% mod) 
R = Edata - HatB %*% t(mod)

R.pc = getPcaResult(R, varNames = colnames(R))
R.pv =  data.frame(Var1 = "Origin", Var2 = rownames(R), value = R.pc$varDf[,2])
# use permutation test to find the unusual large eigenvalues --------------
B = 1000
pvMat = matrix(nrow = B, ncol = dim(R)[1])
tempR = R
for(k in 1:B){
  # make permutation of each row independently
  newE = matrix(nrow = dim(R)[1],ncol = dim(R)[2])
  for(i in 1:dim(R)[1]){
    rank = sample(dim(R)[2])
    newE[i,1:dim(R)[2]] = tempR[i,rank]
  }
  # refit the model to get the new residual matrix
  newR = newE - newE %*% mod %*% solve(t(mod) %*% mod) %*% t(mod)
  
  # do pca on newR and take out the proprotion variance vector
  newR.pc = getPcaResult(newR, varNames = colnames(newR))
  pvMat[k, 1:dim(R)[1]] = newR.pc$varDf[,2]
  temoR = newR
}

colnames(pvMat) = rownames(R)
rownames(pvMat) = paste0("Run", 1:B)

EpvMat = melt(pvMat)

gg1 = ggplot(data = EpvMat, aes(x = Var2, y = value)) + 
  geom_boxplot(col = "blue") +
  geom_point(data = R.pv, aes(x = Var2, y = value), col = "red")

gg2 = ggplot(data = EpvMat, aes(x = Var2, y = value, group = Var1)) + 
  geom_line(col = "blue") +
  geom_line(data = R.pv, aes(x = Var2, y = value), col = "red") +
  geom_point(data = R.pv, aes(x = Var2, y = value), col = "red", size = 2)
print(gg1)
print(gg2)
```


Next we construct SV.
```{r, echo = FALSE, warning=FALSE, message = FALSE}
library(ggplot2)
library(reshape2)
library(sva)
# estimate the surrogate variables
svobj100d = sva(Edata100d,mod_100d,mod0_100d,n.sv= n.sv_be_100d, B = 5)
fsvobj100d = fsva(Edata100d, mod_100d, svobj100d, Edata100d)

# get the data adjusted by the sv
Edb100d = fsvobj100d$db
Mdb100d = melt(Edb100d)
g5.1 = ggplot(Mdb100d, aes(x = Var2, y = Var1, fill = value)) +
  labs(x = "Sample", y = "Gene", fill = "Value",title = "1 Surrogate Variables") +
  geom_tile() + 
  scale_fill_gradient2()
print(g5.1)

Mdb100d.0 = melt(Edb100d[1:10,])
g5.2 = ggplot(Mdb100d.0, aes(x = Var2, y = Var1, fill = value)) +
  labs(x = "Sample", y = "Gene", fill = "Value",title = "1 Surrogate Variables") +
  geom_tile() + 
  scale_fill_gradient2()
print(g5.2)
```

##### Scale different mixture batch effects
There are 80 samples. The samples belong to two classes. First 40 samples belong to Class 1 and others belong to Class 2. The samples are collected from two Batches. Batch 1 contains Sample 1 to Sample 20 and Sample 41 to Sample 60. Other samples come from Batch 2.

Each sample has 100 features.
- $X_1$ is the primary variables. For Class 1, $X_1 = 2$; for Class 2, $X_1 = -2$.
- $X_2$ is the batch effects. For Batch 1, $X_2 \sim \mathrm{N}(0, 0.1)$; for Batch 2, $X_2 \sim \mathrm{N}(0, 2)$.
- $X_3, \dots, X_100$ is the random noise.

```{r, echo = FALSE, warning=FALSE, message = FALSE}
library(ggplot2)
library(reshape2)
library(dplyr)
load("/Users/meileijiang/researchspace/robust-against-heterogeneity/Surrogate Variable Analysis/rdata/scaleToyexample.RData")
M <- train %>% select(-y, -batch)
# Expression data
Edata = t(M)
Mdata = melt(Edata)

g6 = ggplot(Mdata, aes(x = Var2, y = Var1, fill = value)) +
  labs(x = "Sample", y = "Gene", fill = "Value") +
  geom_tile() + 
  scale_fill_gradient2()
print(g6)

Mdata.0 = melt(Edata[1:10, ])
g6.0 = ggplot(Mdata.0, aes(x = Var2, y = Var1, fill = value)) +
  labs(x = "Sample", y = "Gene", fill = "Value") +
  geom_tile() + 
  scale_fill_gradient2()
print(g6.0)
```

Then, we apply sva algorithm to detect the number of surrogate variables.
```{r, echo = FALSE, warning=FALSE, message = FALSE}
library(sva)
# make model matrix -------------------------------------------------------
## full model 
mod = model.matrix(~ y, data = train)
## null model
mod0 = model.matrix(~ 1, data = train)

# estimate the number of latent factors that need to be estimated ---------
n.sv_be = num.sv(Edata,mod,method="be")
print(n.sv_be)
n.sv_leek = num.sv(Edata,mod,method="leek")
print(n.sv_leek)
```

The SVA fails to detect the unmodeled factor. However, I can still try to construct the SV by setting the number of SV as 1 mannually.
```{r, echo = FALSE, warning=FALSE, message = FALSE}
library(ggplot2)
library(reshape2)
library(sva)
# estimate the surrogate variables
svobj = sva(Edata,mod,mod0,n.sv= 1, B = 5)
fsvobj = fsva(Edata, mod, svobj, Edata)

# get the data adjusted by the sv
Edb = fsvobj$db
Mdb = melt(Edb)
g6.1 = ggplot(Mdb, aes(x = Var2, y = Var1, fill = value)) +
  labs(x = "Sample", y = "Gene", fill = "Value",title = "1 Surrogate Variables") +
  geom_tile() + 
  scale_fill_gradient2()
print(g6.1)

Mdb.0 = melt(Edb[1:10,])
g6.2 = ggplot(Mdb.0, aes(x = Var2, y = Var1, fill = value)) +
  labs(x = "Sample", y = "Gene", fill = "Value",title = "1 Surrogate Variables") +
  geom_tile() + 
  scale_fill_gradient2()
print(g6.2)
```

#### Unbalanced Batch Effect
In these simulations, there are 80 samples. The samples belong to two classes. First 40 samples belong to Class 1 and others belong to Class 2. The samples are collected from two batches. But the propotion of samples in Class 1 from Batch 1 is different with the propotionof samples in Class 2 from Batch 1.

##### Mean Heterogeneity Batch Effect

Each sample has 100 features:

- $X_1$ is the primary variables. For Class 1, $X_1 = 2$; for Class 2, $X_1 = -2$.
- $X_2$ is the batch effects. For Batch 1, $X_2 \sim \mathrm{N}(2, 1)$; for Batch 2, $X_2 \sim \mathrm{N}(-2, 1)$. $\pi_1$ of Class 1 from Batch 1 and $\pi_2$ of Class 2 from Batch 1. The others from Batch 2.
- $X_3, \dots, X_100$ is the random noise.

We do simulation for three sets of $\pi_1$ and $\pi_2$:

- $\pi_1 = 0.7, \pi_2 = 0.2$. In this case, the samples from Batch 1 and Batch 2 are different.
- $\pi_1 = 0.33, \pi_2 = 0.67$.
- $\pi_1 = 0.9, \pi_2 = 0.1$.

```{r, echo = FALSE, warning=FALSE, message = FALSE}
library(dplyr)
library(ggplot2)
library(sva)
library(reshape2)
library(gridExtra)
# load the data and functions
load("/Users/meileijiang/researchspace/robust-against-heterogeneity/Surrogate Variable Analysis/rdata/unbalanced_mean_hetero_example.RData")
source('~/researchspace/robust-against-heterogeneity/Surrogate Variable Analysis/pcafuns/R/pcaScreePlot.R')
source('~/researchspace/robust-against-heterogeneity/Surrogate Variable Analysis/pcafuns/R/getPcaResult.R')

# data processing
M1 <- train %>% select(-y, -batch)
M2 <- train2%>% select(-y, -batch)
M3 <- train3%>% select(-y, -batch)
## Expression data
Edata1 = t(M1)
colnames(Edata1) = paste0("sample", c(1: dim(Edata1)[2]))
Mdata1 = melt(Edata1)

Edata2 = t(M2)
colnames(Edata2) = paste0("sample", c(1: dim(Edata2)[2]))
Mdata2 = melt(Edata2)

Edata3 = t(M3)
colnames(Edata3) = paste0("sample", c(1: dim(Edata3)[2]))
Mdata3 = melt(Edata3)

Y <- train %>% select(y)

# make model matrix -------------------------------------------------------
## full model 
mod = model.matrix(~ y, data = train)
## null model
mod0 = model.matrix(~ 1, data = train)


# estimate the number of latent factors that need to be estimated ---------
n.sv1 = num.sv(Edata1,mod,method="be")
n.sv2 = num.sv(Edata2,mod,method="be")
n.sv3 = num.sv(Edata3,mod,method="be")

print(paste0("n.sv1 = ",n.sv1,". Fail to find out the batch effect."))
print(paste0("n.sv2 = ",n.sv2,". Fail to Find out the batch effect."))
print(paste0("n.sv3 = ",n.sv3,". Fail to find out the batch effect."))
# it fails to estimate the number of surogate variable

print("Manually set the number of surrogate variable as 1 and apply sva to remove the effects of surrogate variable.")
# estimate the surrogate variables
svobj1 = sva(t(M1),mod,mod0,n.sv= 1, B = 5)
svobj2 = sva(t(M2),mod,mod0,n.sv= 1, B = 5)
svobj3 = sva(t(M3),mod,mod0,n.sv= 1, B = 5)

# Remove the batch effects
fsvobj1 = fsva(Edata1, mod, svobj1, Edata1)
fsvobj2 = fsva(Edata2, mod, svobj2, Edata2)
fsvobj3 = fsva(Edata3, mod, svobj3, Edata3)

Edb1 = fsvobj1$db
Edb2 = fsvobj2$db
Edb3 = fsvobj3$db

trainSV1 <- data.frame(Y, t(Edb1), batch = train$batch)
trainSV2 <- data.frame(Y, t(Edb2), batch = train2$batch)
trainSV3 <- data.frame(Y, t(Edb3), batch = train3$batch)

gg1 = ggplot(data = train, aes(x = as.factor(batch), y = X2)) + 
  geom_boxplot() +
    ylim(-5, 5) + 
  labs(title = 'Before adjusting \n
pi_1 = 0.7, pi_2 = 0.2', x = "Batch")

gg2 = ggplot(data = train2, aes(x = as.factor(batch), y = X2)) + 
  geom_boxplot() +
  ylim(-5, 5) + 
  labs(title = 'Before adjusting \n
pi_1 = 0.33, pi_2 = 0.67', x = "Batch")

gg3 = ggplot(data = train3, aes(x = as.factor(batch), y = X2)) + 
  geom_boxplot() +
  ylim(-5, 5) + 
  labs(title = 'Before adjusting \n
pi_1 = 0.9, pi_2 = 0.1', x = "Batch")

grid.arrange(gg1,gg2,gg3, nrow = 1, ncol = 3)

gg4 = ggplot(data = trainSV1, aes(x = as.factor(batch), y = X2)) + 
  geom_boxplot() +
  ylim(-5, 5) + 
  labs(title = 'After adjusting  \n
pi_1 = 0.7, pi_2 = 0.2', x = "Batch")

gg5 = ggplot(data = trainSV2, aes(x = as.factor(batch), y = X2)) + 
  geom_boxplot() +
  ylim(-5, 5) + 
  labs(title = 'After adjusting  \n
pi_1 = 0.33, pi_2 = 0.67', x = "Batch")

gg6 = ggplot(data = trainSV3, aes(x = as.factor(batch), y = X2)) + 
  geom_boxplot() +
  ylim(-5, 5) + 
  labs(title = 'After adjusting  \n
pi_1 = 0.9, pi_2 = 0.1', x = "Batch")

grid.arrange(gg4,gg5,gg6, nrow = 1, ncol = 3)

# dig into the analysis ---------------------------------------------------
print("Analysis of detecting the surgoate variables")
# estimate the coefficient of basis matrix
R1 = Edata1 - Edata1 %*% mod %*% solve(t(mod) %*% mod)  %*% t(mod)
R2 = Edata2 - Edata2 %*% mod %*% solve(t(mod) %*% mod)  %*% t(mod)
R3 = Edata3 - Edata3 %*% mod %*% solve(t(mod) %*% mod)  %*% t(mod)

R.pc1 = getPcaResult(R1, varNames = colnames(R1))
R.pc2 = getPcaResult(R2, varNames = colnames(R2))
R.pc3 = getPcaResult(R3, varNames = colnames(R3))

R.pv1 =  data.frame(Var1 = "Origin", Var2 = rownames(R.pc1$varDf), value = R.pc1$varDf[,2])
R.pv2 =  data.frame(Var1 = "Origin", Var2 = rownames(R.pc2$varDf), value = R.pc2$varDf[,2])
R.pv3 =  data.frame(Var1 = "Origin", Var2 = rownames(R.pc3$varDf), value = R.pc3$varDf[,2])

# use permutation test to find the unusual large eigenvalues --------------
B = 1000
pvMat1 = matrix(nrow = B, ncol = dim(R.pc1$varDf)[1])
pvMat2 = matrix(nrow = B, ncol = dim(R.pc2$varDf)[1])
pvMat3 = matrix(nrow = B, ncol = dim(R.pc3$varDf)[1])

tempR = R1
for(k in 1:B){
  # make permutation of each row independently
  newE = matrix(nrow = dim(R.pc1$varDf)[1],ncol = dim(R1)[2])
  for(i in 1:dim(R.pc1$varDf)[1]){
    rank = sample(dim(R1)[2])
    newE[i,1:dim(R1)[2]] = tempR[i,rank]
  }
  # refit the model to get the new residual matrix
  newR = newE - newE %*% mod %*% solve(t(mod) %*% mod) %*% t(mod)
  
  # do pca on newR and take out the proprotion variance vector
  newR.pc = getPcaResult(newR, varNames = colnames(newR))
  pvMat1[k, 1:dim(R.pc1$varDf)[1]] = newR.pc$varDf[,2]
  tempR = newR
}

tempR = R2
for(k in 1:B){
  # make permutation of each row independently
  newE = matrix(nrow = dim(R.pc2$varDf)[1],ncol = dim(R2)[2])
  for(i in 1:dim(R.pc2$varDf)[1]){
    rank = sample(dim(R2)[2])
    newE[i,1:dim(R2)[2]] = tempR[i,rank]
  }
  # refit the model to get the new residual matrix
  newR = newE - newE %*% mod %*% solve(t(mod) %*% mod) %*% t(mod)
  
  # do pca on newR and take out the proprotion variance vector
  newR.pc = getPcaResult(newR, varNames = colnames(newR))
  pvMat2[k, 1:dim(R.pc2$varDf)[1]] = newR.pc$varDf[,2]
  tempR = newR
}

tempR = R3
for(k in 1:B){
  # make permutation of each row independently
  newE = matrix(nrow = dim(R.pc3$varDf)[1],ncol = dim(R3)[2])
  for(i in 1:dim(R.pc1$varDf)[1]){
    rank = sample(dim(R3)[2])
    newE[i,1:dim(R3)[2]] = tempR[i,rank]
  }
  # refit the model to get the new residual matrix
  newR = newE - newE %*% mod %*% solve(t(mod) %*% mod) %*% t(mod)
  
  # do pca on newR and take out the proprotion variance vector
  newR.pc = getPcaResult(newR, varNames = colnames(newR))
  pvMat3[k, 1:dim(R.pc3$varDf)[1]] = newR.pc$varDf[,2]
  tempR = newR
}

colnames(pvMat1) = rownames(R.pc1$varDf)
rownames(pvMat1) = paste0("Run", 1:B)
colnames(pvMat2) = rownames(R.pc2$varDf)
rownames(pvMat2) = paste0("Run", 1:B)
colnames(pvMat3) = rownames(R.pc3$varDf)
rownames(pvMat3) = paste0("Run", 1:B)

EpvMat1 = melt(pvMat1)
pv_stat1 = EpvMat1 %>% 
  group_by(Var2) %>%
  summarise(Median = median(value), Q975 = quantile(value, .975), Q025 = quantile(value, .025))

EpvMat2 = melt(pvMat2)
pv_stat2 = EpvMat2 %>% 
  group_by(Var2) %>%
  summarise(Median = median(value), Q975 = quantile(value, .975), Q025 = quantile(value, .025)) 

EpvMat3 = melt(pvMat3)
pv_stat3 = EpvMat3 %>% 
  group_by(Var2) %>%
  summarise(Median = median(value), Q975 = quantile(value, .975), Q025 = quantile(value, .025)) 

gg7 = ggplot(data = pv_stat1, aes(x = Var2, y = Median) )+ 
  geom_point(col = "blue") +
  geom_errorbar(aes(ymax = Q975, ymin = Q025)) + 
  geom_point(data = R.pv1, aes(x = Var2, y = value), col = "red")

gg8 = ggplot(data = EpvMat1, aes(x = Var2, y = value, group = Var1)) + 
  geom_line(col = "blue") +
  geom_line(data = R.pv1, aes(x = Var2, y = value), col = "red") +
  geom_point(data = R.pv1, aes(x = Var2, y = value), col = "red", size = 2)

grid.arrange(gg7,gg8, nrow = 1, ncol = 2)

gg9 = ggplot(data = pv_stat2, aes(x = Var2, y = Median) )+ 
  geom_point(col = "blue") +
  geom_errorbar(aes(ymax = Q975, ymin = Q025)) + 
  geom_point(data = R.pv2, aes(x = Var2, y = value), col = "red")

gg10 = ggplot(data = EpvMat2, aes(x = Var2, y = value, group = Var1)) + 
  geom_line(col = "blue") +
  geom_line(data = R.pv2, aes(x = Var2, y = value), col = "red") +
  geom_point(data = R.pv2, aes(x = Var2, y = value), col = "red", size = 2)

grid.arrange(gg9, gg10, nrow = 1, ncol = 2)

gg11 = ggplot(data = pv_stat3, aes(x = Var2, y = Median) )+ 
  geom_point(col = "blue") +
  geom_errorbar(aes(ymax = Q975, ymin = Q025)) + 
  geom_point(data = R.pv3, aes(x = Var2, y = value), col = "red")

gg12 = ggplot(data = EpvMat3, aes(x = Var2, y = value, group = Var1)) + 
  geom_line(col = "blue") +
  geom_line(data = R.pv3, aes(x = Var2, y = value), col = "red") +
  geom_point(data = R.pv3, aes(x = Var2, y = value), col = "red", size = 2)

grid.arrange(gg11, gg12, nrow = 1, ncol = 2)

```